# Exercise 3

## Recap

In the first and second R sessions, we covered the following:

::::::: row
:::::: col-sm-6
::::: {.card .border-primary .mb-3 style="max-width: 30rem;"}
::: card-header
Exercise 1 and 2 Tasks
:::

::: {.card-body .text-primary}
<p class="card-text">

T1: Downloading and Loading ESS Subset on R. ✅<br> 
T2: Introduction to Key Commands ✅ <br> 
T3: Plotting Variable Distributions ✅<br> 
T4: Cross Country and Mode Comparision ✅<br>
T5: Splitting Variable of Choice ✅ 
</p>
:::
:::::
::::::
:::::::

# Recap Exercise

For this exercise, we have some R cheat sheets and a BINGO! But first, a recap exercise, so we do not forget what we did in the last two sessions!  

- Load the ESS subset dataset into your environment  
- Confirm that the dataset is available  
- View the dataset  
- Check its dimensions  
- Look at the column names  
- Explore any two variable using basic summaries  
- Create a quick distribution plot for any variable 
- Compare political interest across countries; look at both counts and proportions  
- Choose another variable and split it into two groups
- Compare another variable across these groups  


## Cheat Sheets and Bingo!

These cheat sheets help list various data transformations and tidying your environment code syntax without needing to memorize every single function argument:

* [Data Transformation with dplyr](https://opensource.posit.co/resources/cheatsheets/data-transformation/)
* [Data Tidying with tidyr](https://opensource.posit.co/resources/cheatsheets/tidyr/)

## Splitting variable (``mutate() & cut()``)

::: {.panel-tabset}

### Piping + Operators

From [Pipe in R: A Guide](https://builtin.com/articles/pipe-in-r#:~:text=The%20pipe%20operator%2C%20formerly%20written,a%20sequence%20of%20analysis%20steps.)

The pipe operator, formerly written as %>%, was a longstanding feature of the magrittr package for R. The new version, written as |>, is largely the same. It takes the output of one function and passes it into another function as an argument. This allows us to link a sequence of analysis steps.

To visualize this process, imagine a factory with different machines placed along a conveyor belt. Each machine is a function that performs a stage of our analysis, like filtering or transforming data. The pipe works like a conveyor belt, transporting the output of one machine to another for further processing.

The pipe is best understood as the word **"then"**, allowing us to read code like a sentence. We will use it here to select our specific variables and filter out a country.

To look at how it works be can check an example from the ``ESS_Subset``, using a popular set of commands. 

```{r,echo=F, message=FALSE}
library(readr)
ESS_Subset <- read_csv("ESS_Subset.csv")
```


```{r, warning=F, message=FALSE}
library(tidyverse)
DE_Subset <- ESS_Subset %>%
  select(cntry, agea, gndr, eduyrs, polintr, lrscale) %>%
  filter(cntry == "DE") 
```



### Use case + Documentation


### When 

###  How

### ESS Example

```{r}
ESS_Subset <- ESS_Subset %>%
  mutate(gender_label = case_when(
    gndr == 1 ~ "Male",
    gndr == 2 ~ "Female",
    TRUE      ~ NA_character_  
  ))
```


```{r}
ESS_Subset <- ESS_Subset %>%
  mutate(age_group = cut(agea, 
                         breaks = c(0, 30, 60, 120), 
                         labels = c("Young", "Middle", "Senior")))
```



:::

## Sub-populations (``filter()``)

::: {.panel-tabset}

### Use case + Documentation


### When 


###  How

### ESS Example


```{r}
#Simple ESS Example
# Filtering only Germany

df_de <- ESS_Subset %>% filter(cntry == "DE")
```

```{r}
# Nuanced ESS examples
# Filtering young people who are highly interested in politics

df_active_youth <- ESS_Subset %>% 
  filter(agea < 30, polintr <= 2)

```

When we mant to filter out multiple specific values without writing ten ``==``statements

```{r}
# Filter for 2  countries of interest
target_countries <- c("DE","SE")
ESS_2country <- ESS_Subset %>% 
  filter(cntry %in% target_countries)

# Filtering for specific countries AND valid answers (Excluding 7, 8, and 9 from the polintr variable)
df_clean <- ESS_Subset %>%
  filter(cntry %in% c("DE", "SE"),
         !polintr %in% c(7, 8, 9))
```

:::


## Combining Datasets (``merge()``)

::: {.panel-tabset}

### Use case + Documentation


### When 


###  How

### ESS Example

```{r}
#Simple ESS Example
# Sometimes it is possible to merge two datasets as well! In this example let us add the full names of the country using merge

# Note: This is a simple merge example, it is better to use mutate when basically adding new columns like this! 

# First check what are the countries we have
table(ESS_Subset$cntry)

# Next, make a df we want to merge, many times we already have it! 
country_names <- tibble(cntry = c("DE", "HU", "SE"), full_name = c("Germany", "Hungary", "Sweden"))


df_joined <- ESS_Subset %>% left_join(country_names, by = "cntry")

```


```{r}
# Nuanced ESS example
# What if there was another dataset we can merge this data with! 
# *This is just an example

library(gapminder)

# gapminder has full country names, so we also add those!
merge_help <- tibble(cntry = c("DE", "HU", "SE"), 
                 country = c("Germany", "Hungary", "Sweden"))

ess_with_names <- ESS_Subset %>% left_join(merge_help, by = "cntry")

# since we do not want the combinated dataset to explode, lets filter for only one year, in this data the latest is 2007. 
gap_filtered <- gapminder %>% filter(year == 2007)
combined_data <- ess_with_names %>% left_join(gap_filtered, by = "country")

```

:::


## Aggregations (``group_by()``)


::: {.panel-tabset}


### Use case + Documentation


### When 


###  How

### ESS Example

```{r}
#Simple ESS Example
ESS_Subset %>% 
  group_by(gndr) %>% 
  summarize(mean_age = mean(agea, na.rm = TRUE))

```

```{r}
# Nuanced ESS examples

trust_summary <- ESS_Subset %>%
  group_by(cntry) %>%
  summarize(
    avg_trust_police = mean(trstplc, na.rm = TRUE),
    respondents = n()
  )
```

:::

## Reshaping data (``pivot_long() & pivot_wider()``)

::: {.panel-tabset}


### Use case + Documentation

```{r}
# Let us understand this using gapminder. 
gap_wide <- gapminder %>%
  filter(country %in% c("Germany", "Hungary", "Sweden"), 
         year %in% c(1997, 2002, 2007)) %>%
  select(country, year, gdpPercap) %>%
  pivot_wider(names_from = year, values_from = gdpPercap)

# helpful for ggplot!
gap_long <- gap_wide %>%
  pivot_longer(cols = `1997`:`2007`, 
               names_to = "year", 
               values_to = "gdp")

```


### When 


###  How

### ESS Example

```{r}
# Without Pivot
trust_comparison_nopiv <- ESS_Subset %>%
  group_by(cntry) %>%
  summarize(
    Police = mean(trstplc, na.rm = TRUE),
    Parliament = mean(trstprl, na.rm = TRUE)
  ) 

# With Pivot
# Pivoting helps with making cleaner tables and ggplot graphs! 
trust_comparison_piv <- ESS_Subset %>%
  group_by(cntry) %>%
  summarize(
    Police = mean(trstplc, na.rm = TRUE),
    Parliament = mean(trstprl, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(Police, Parliament), 
    names_to = "Institution", 
    values_to = "Avg_Trust_Score"
  )

```



:::


## Transformations (``a  <- b +c``)

::: {.panel-tabset}


### Use case + Documentation


### When 


###  How

### ESS Example
```{r}
#Simple ESS Example
ESS_Subset <- ESS_Subset %>% 
  mutate(total_society = trstplc + trstprl)



```

```{r}
# Nuanced ESS examples
ESS_Subset <- ESS_Subset %>%
  filter(imwbcnt <= 10, 
         imueclt <= 10, 
         imbgeco <= 10) %>% 
  mutate(immig_index = rowMeans(
    select(., imwbcnt, imueclt, imbgeco), 
    na.rm = FALSE 
  ))
```


:::


## Regressions (``lm() & glm()``)

::: {.panel-tabset}


### Use case + Documentation


### When 


###  How

### ESS Example

:::


## Full Analysis Pipeline


